---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# GXwasR

A tool for conducting sex-aware quality control, association analysis, and testing various models of sex-dependent genetic effects in complex traits.

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
<!-- badges: end -->

This package implements various statistical genetics models for Genome-Wide Association (GWA) and X-Chromosome Wide Association (XWA) analyses in a sex-combined or sex-stratified way considering X-Chromosome Inactivation (XCI) pattern. In addition to association analysis, the package also enables testing for sex differences in genetic effects, including the implementation of specific models and applying best practices for additional quality control (QC) of genetic data required for these tests. The package includes twenty-five different functions in six different categories (A-F) which enable a comprehensive pipeline for sex-aware genetic association analysis of common variants with unrelated individuals.

**(A)Pre-imputation QC:**  `QCsnp()`; `QCsample()`; `AncestryCheck()`; `SexCheck()`

**(B)Post-imputation QC:** `QCsnp()`; `QCsample2()`; `Xhwe()`; `MAFdiffSexControl()`; `FilterRegion()`

**(C)Sex-combined and sex-stratified GWAS with XWAS:** `GXwasR()`

**(D)Sex-differential test:** `SexDiff()`; `SexDiffZscore()`; `DiffZeroOne()`

**(E)High level analysis:** `TestXGene()`; `MetaGWAS()`; `ComputePRS()`; `ComputeCorrBT()`; `EstimateHerit()`

**(F)Utility Functions:** `FilterPlinkSample()`; `ComputeGeneticPC()`; `ClumpLD()`; `GetMFPlink()`; `plinkVCF()`; `MergeRegion()`; `FilterAllele()`; `PlinkSummary()`

## 📦 Installation

You can install the development version of GXwasR from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("boseb/GXwasR")
```

## External Dependency: PLINK

This package depends on the [PLINK](https://www.cog-genomics.org/plink/) command-line tool (version 1.9 or 2.0). PLINK must be installed separately and made available on your system.

PLINK is not bundled with this package and must either:

- be on your system PATH, or
- be specified via the PLINK_PATH environment variable.

### 🔧 PLINK Installation Instructions

Binaries for all major platforms can be downloaded from:

- [PLINK v1.9](https://www.cog-genomics.org/plink/1.9/)
- [PLINK v2.0](https://www.cog-genomics.org/plink/2.0/)

Detailed, platform-specific setup instructions can be found in the INSTALL file included with this package.

### 🧭 Configuring the PLINK Path

This package will attempt to locate PLINK using:

- The `PLINK_PATH` environment variable, if set.
- The system path, via `Sys.which("plink")`.

If PLINK is not found, an error will be raised with guidance on how to resolve it.

You can manually set the path in your R session:

```r
Sys.setenv(PLINK_PATH = "/path/to/plink")
```

For a persistent configuration, you can add this line to your .Renviron file:

```
PLINK_PATH=/path/to/plink
```

To verify that PLINK is discoverable:

```r
plink_path <- Sys.getenv("PLINK_PATH", unset = Sys.which("plink"))
if (!file.exists(plink_path) || !nzchar(plink_path)) {
  stop("PLINK binary not found. Please install PLINK and/or set the PLINK_PATH environment variable.")
}
```

## External Dependency: GCTA

This package also utilizes the [GCTA](https://yanglab.westlake.edu.cn/software/gcta/#Overview) command-line tool (Genome-wide Complex Trait Analysis). GCTA must be installed separately and made available on your system.

GCTA is not bundled with this package and must either:

-be on your system PATH, or
- be specified via the GCTA_PATH environment variable.

### 🔧 GCTA Installation Instructions

Binaries for all major platforms can be downloaded from the [GCTA](https://yanglab.westlake.edu.cn/software/gcta/#Overview) website.

Detailed, platform-specific setup instructions can be found in the INSTALL file included with this package.

### 🧭 Configuring the GCTA Path

This package will attempt to locate GCTA using:

- The GCTA_PATH environment variable, if set.
- The system path, via Sys.which("gcta64").

If GCTA is not found, an error will be raised with guidance on how to resolve it.

You can manually set the path in your R session:

```r
Sys.setenv(GCTA_PATH = "/path/to/gcta64")
```

For a persistent configuration, you can add this line to your .Renviron file:

```
GCTA_PATH=/path/to/gcta64
```

To verify that GCTA is discoverable:

```r
gcta_path <- Sys.getenv("GCTA_PATH", unset = Sys.which("gcta64"))
if (!file.exists(gcta_path) || !nzchar(gcta_path)) {
  stop("GCTA binary not found. Please install GCTA and/or set the GCTA_PATH environment variable.")
}
```

## Example Analysis

Run a genome-wide association study (GWAS) and X-chromosome-wide association study (XWAS):

```{r example}
library(GXwasR)
ResultGXwas <- GXwas(
  DataDir = system.file("extdata", package = "GXwasR"),
  ResultDir = tempdir(), 
  finput = "GXwasR_example", 
  trait = "binary", 
  xmodel = "FMstatrified", 
  covarfile = NULL, 
  sex = TRUE, 
  xsex = FALSE, 
  combtest = "fisher.method", 
  MF.p.corr = "none",
  snp_pval = 1e-08, 
  plot.jpeg = FALSE, 
  suggestiveline = 5, 
  genomewideline = 7.3,
  MF.mc.cores = 1, 
  ncores = 0
  )
ResultGXwas
```

## Code of Conduct

Please note that the GXwasR project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
