---
title: "Tutorial for heritability estimate and computing genetic correlation in GXwasR R package."
author: 
  - name: "Banabithi Bose"
    affiliation:
    - University of Colorado Anschutz Medical; 
    - Northwestern University
    email: banabithi.bose@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: false
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('GXwasR')`"
vignette: >
  %\VignetteIndexEntry{GXwasR_heritability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    fig.path = "./figures/",
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Bib setup
library("RefManageR")

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    GXwasR = citation("GXwasR")[1]
)
```

**This document provides a tutorial for heritability estimate and computing genetic correlation utilizing GXwasR R package.**

```{r}
## Call some libraries
library(GXwasR)
library(printr)
library(rmarkdown)
```
# <a id = "EstimateHerit():"> </a>Computing heritability estimate using EstimateHerit():

```{r}
help(EstimateHerit, package = GXwasR)
```

## <a id = "Running EstimateHerit() with model = 'GREML'"> </a> Running EstimateHerit() with model = "GREML"

```{r}
# Running
data("highLD_hg19", package = "GXwasR")
DataDir <- GXwasR:::GXwasR_data()
ResultDir = tempdir()
finput <- "GXwasR_example"
ncores = 3
model = "GREML"
byCHR = FALSE
r2_LD = 0
LDSC_blocks = 20
REMLalgo = 0
nitr = 100
cat_covarfile = NULL
quant_covarfile = NULL
prevalence = 0.01
partGRM = FALSE
autosome = TRUE
Xsome = TRUE
nGRM = 3
cripticut = 0.025
minMAF = NULL
maxMAF = NULL
hg = "hg19"
byCHR = TRUE
PlotIndepSNP = TRUE
IndepSNP_window_size = 50
IndepSNP_step_size = 5
IndepSNP_r2_threshold = 0.02

# For saving the plot in .jpeg file, make plotjpeg = TRUE.

H2 <- EstimateHerit(DataDir = DataDir, ResultDir = ResultDir, finput = finput, summarystat = NULL, ncores = 5, model = model, byCHR = TRUE, r2_LD = 0, LDSC_blocks = 20,REMLalgo = 0, nitr = nitr, cat_covarfile = NULL, quant_covarfile = NULL,prevalence = 0.01, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3, computeGRM = TRUE, grmfile_name = NULL, cripticut = 0.025, minMAF = NULL, maxMAF = NULL,hg = "hg19",PlotIndepSNP = TRUE, IndepSNP_window_size = 50,IndepSNP_step_size = 5,IndepSNP_r2_threshold = 0.02,highLD_regions = highLD_hg19, plotjpeg = TRUE, plotname = "Heritability_Plots")

#Heritability
H2
```

## Running the EstimateHerit() by providing chromosome-wise GRM files in ResultDir
Set:

`computeGRM` = FALSE

`grmfile_name` = "GXwasR"...... This is the prefix. These files are created in the previous run in ResultDir. Users are encouraged to check that folder.


```{r}

H2 <- EstimateHerit(DataDir = DataDir, ResultDir = ResultDir, finput = finput, summarystat = NULL, ncores = 5, model = model, byCHR = TRUE, r2_LD = 0, LDSC_blocks = 20,REMLalgo = 0, nitr = nitr, cat_covarfile = NULL, quant_covarfile = NULL,prevalence = 0.01, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3, computeGRM = FALSE, grmfile_name = "GXwasR", cripticut = 0.025, minMAF = NULL, maxMAF = NULL,hg = "hg19",PlotIndepSNP = TRUE, IndepSNP_window_size = 50,IndepSNP_step_size = 5,IndepSNP_r2_threshold = 0.02,highLD_regions = highLD_hg19, plotjpeg = FALSE, plotname = "Heritability_Plots")

#Heritability
H2
```

## Running the EstimateHerit() for genome-wide heritability estimate by creating GRM files
Set:

computeGRM = TRUE

byCHR = FALSE

grmfile_name = NULL.......if computeGRM = TRUE the grmfile_name is by default NULL.

In genome-wide case, there no plot will be generated.

```{r}

H2 <- EstimateHerit(DataDir = DataDir, ResultDir = ResultDir, finput = finput, summarystat = NULL, ncores = 5, model = model, byCHR = FALSE, r2_LD = 0, LDSC_blocks = 20,REMLalgo = 0, nitr = nitr, cat_covarfile = NULL, quant_covarfile = NULL,prevalence = 0.01, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3, computeGRM = TRUE, grmfile_name = NULL, cripticut = 0.025, minMAF = NULL, maxMAF = NULL,hg = "hg19",PlotIndepSNP = TRUE, IndepSNP_window_size = 50,IndepSNP_step_size = 5,IndepSNP_r2_threshold = 0.02,highLD_regions = highLD_hg19, plotjpeg = FALSE, plotname = "Heritability_Plots")

#Heritability
H2
```

## Running the EstimateHerit() for genome-wide heritability estimate by utilizing the pre-computed GRM files in ResultDir
Set:

computeGRM = FALSE

byCHR = FALSE

grmfile_name = "GXwasR"

In genome-wide case, there no plot will be generated.

```{r}

H2 <- EstimateHerit(DataDir = DataDir, ResultDir = ResultDir, finput = finput, summarystat = NULL, ncores = 5, model = model, byCHR = FALSE, r2_LD = 0, LDSC_blocks = 20,REMLalgo = 0, nitr = nitr, cat_covarfile = NULL, quant_covarfile = NULL,prevalence = 0.01, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3, computeGRM = FALSE, grmfile_name = "GXwasR", cripticut = 0.025, minMAF = NULL, maxMAF = NULL,hg = "hg19",PlotIndepSNP = TRUE, IndepSNP_window_size = 50,IndepSNP_step_size = 5,IndepSNP_r2_threshold = 0.02,highLD_regions = highLD_hg19, plotjpeg = FALSE, plotname = "Heritability_Plots")

#Heritability
H2
```

## <a id = "Running EstimateHerit() with model = 'LDSC'"> </a> Running EstimateHerit() genome-wide with model = "LDSC" 
```{r}
data("Summary_Stat_Ex1", package = "GXwasR")
model <- "LDSC"
data("GXwasRData")
DataDir <- GXwasR:::GXwasR_data()
ResultDir = tempdir()
finput <- "GXwasR_example"
test.sumstats <- na.omit(Summary_Stat_Ex1[Summary_Stat_Ex1$TEST=="ADD",c(1:4,6:8)])
colnames(test.sumstats) <- c("chr","rsid","pos","a1","n_eff","beta","beta_se")
summarystat = test.sumstats
highLD_regions = highLD_hg19

H2 <- EstimateHerit(DataDir = DataDir, ResultDir = ResultDir, finput = "GXwasR_example", summarystat = summarystat, ncores, model = model, byCHR = FALSE, r2_LD = 0, LDSC_blocks = 20,REMLalgo = 0, nitr = 100, cat_covarfile = NULL, quant_covarfile = NULL,prevalence = 0.01, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3,cripticut = 0.025, minMAF = NULL, maxMAF = NULL,hg = "hg19",PlotIndepSNP = TRUE, IndepSNP_window_size = 50,IndepSNP_step_size = 5,IndepSNP_r2_threshold = 0.02,highLD_regions = highLD_hg19)
#Heritability
H2
```

## <a id = "Running EstimateHerit() with model = 'LDSC'"> </a> Running EstimateHerit() chromosome-wise with model = "LDSC"
```{r}
model <- "LDSC"
DataDir <- GXwasR:::GXwasR_data()
ResultDir = tempdir()
finput <- "GXwasR_example"
test.sumstats <- na.omit(Summary_Stat_Ex1[Summary_Stat_Ex1$TEST=="ADD",c(1:4,6:8)])
colnames(test.sumstats) <- c("chr","rsid","pos","a1","n_eff","beta","beta_se")
summarystat = test.sumstats
highLD_regions = highLD_hg19

H2 <- EstimateHerit(DataDir = DataDir, ResultDir = ResultDir, finput = "GXwasR_example", summarystat = summarystat, ncores, model = model, byCHR = TRUE, r2_LD = 0, LDSC_blocks = 20,REMLalgo = 0, nitr = 100, cat_covarfile = NULL, quant_covarfile = NULL,prevalence = 0.01, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3,cripticut = 0.025, minMAF = NULL, maxMAF = NULL,hg = "hg19",PlotIndepSNP = TRUE, IndepSNP_window_size = 50,IndepSNP_step_size = 5,IndepSNP_r2_threshold = 0.02,highLD_regions = highLD_hg19)

# Chromosome-wise Heritability Estimate
H2
```

# <a id = "GeneticCorrBT()"> </a> Computing genetic correlation between traits using GeneticCorrBT()
```{r}
help(GeneticCorrBT, package = GXwasR)
```
## <a id = "Running GeneticCorrBT()"> </a> Running GeneticCorrBT() by computing GRM genome-wide
```{r}

# Running
DataDir <- GXwasR:::GXwasR_data()
ResultDir <- tempdir()
finput <- "GXwasR_example"
byCHR = FALSE
REMLalgo = 0
nitr <- 3
ncores <- 3
data("Example_phenofile", package = "GXwasR")
phenofile <- Example_phenofile #Cannot be NULL, the interested phenotype column should be labeled as
cat_covarfile <- NULL
quant_covarfile <- NULL
partGRM <- FALSE #Partition the GRM into m parts (by row),
autosome = TRUE
Xsome <- TRUE
cripticut = 0.025
minMAF <- 0.01 # if MAF filter apply
maxMAF <- 0.04
excludeResidual = TRUE
GC <- GeneticCorrBT(DataDir = DataDir, ResultDir = ResultDir, finput = finput, byCHR = byCHR,
                    REMLalgo = 0, nitr = nitr, phenofile = phenofile, cat_covarfile = NULL, quant_covarfile = NULL, computeGRM = TRUE, grmfile_name = NULL, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3,cripticut = 0.025, minMAF = NULL, maxMAF = NULL,excludeResidual = TRUE, ncores = ncores)
##Genetic correlation from the last iteration. In this example there was no rg value since the program stopped after nitr-1 = 2 iteration due to convergence issue.
GC
```

## Running GeneticCorrBT() using precomputed genome-wide GRM 
```{r}

byCHR = FALSE
computeGRM = FALSE
grmfile_name = "GXwasR"

GC <- GeneticCorrBT(DataDir = DataDir, ResultDir = ResultDir, finput = finput, byCHR = byCHR,
                    REMLalgo = 0, nitr = nitr, phenofile = phenofile, cat_covarfile = NULL, quant_covarfile = NULL, computeGRM = FALSE, grmfile_name = grmfile_name, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3,cripticut = 0.025, minMAF = NULL, maxMAF = NULL,excludeResidual = TRUE, ncores = ncores)
##Genetic correlation from the last iteration. In this example there was no rg value since the program stopped after nitr-1 = 2 iteration due to convergence issue.
GC
```

## Running GeneticCorrBT() by computing GRM chromosome-wise
```{r}
byCHR = TRUE
computeGRM = TRUE
grmfile_name = NULL

GC <- GeneticCorrBT(DataDir = DataDir, ResultDir = ResultDir, finput = finput, byCHR = byCHR,
                    REMLalgo = 0, nitr = nitr, phenofile = phenofile, cat_covarfile = NULL, quant_covarfile = NULL, computeGRM = computeGRM, grmfile_name = grmfile_name, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3,cripticut = 0.025, minMAF = NULL, maxMAF = NULL,excludeResidual = TRUE, ncores = ncores)
GC
```

## Running GeneticCorrBT() using chromosome-wide pre-computed GRM
```{r}
byCHR = TRUE
computeGRM = FALSE
grmfile_name = "GXwasR"

GC <- GeneticCorrBT(DataDir = DataDir, ResultDir = ResultDir, finput = finput, byCHR = byCHR,
                    REMLalgo = 0, nitr = nitr, phenofile = phenofile, cat_covarfile = NULL, quant_covarfile = NULL, computeGRM = computeGRM, grmfile_name = grmfile_name, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3,cripticut = 0.025, minMAF = NULL, maxMAF = NULL,excludeResidual = TRUE, ncores = ncores)
GC
```

Note: When computeGRM = FALSE, the function assumes that all required GRM files already exist in ResultDir with the expected naming convention (e.g., Chr1_GXwasR.grm.*). If these files are not present, the analysis will fail because GCTA will be unable to locate the input GRM files.

# Citing `GXwasR`

We hope that `r BiocStyle::Biocpkg("GXwasR")` will be useful for your research. Please use the following information to cite the package and the overall approach. Thank you!

```{r "citation"}
## Citation info
citation("GXwasR")
```

# Reproducibility

The `r BiocStyle::Biocpkg("GXwasR")` package `r Citep(bib[["GXwasR"]])` was made possible thanks to:

* R `r Citep(bib[["R"]])`
* `r BiocStyle::Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
* `r BiocStyle::CRANpkg("knitr")` `r Citep(bib[["knitr"]])`
* `r BiocStyle::CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`
* `r BiocStyle::CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])`
* `r BiocStyle::CRANpkg("sessioninfo")` `r Citep(bib[["sessioninfo"]])`
* `r BiocStyle::CRANpkg("testthat")` `r Citep(bib[["testthat"]])`

This package was developed using `r BiocStyle::Biocpkg("biocthis")`.

`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```

# Bibliography

This vignette was generated using `r BiocStyle::Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
with `r BiocStyle::CRANpkg("knitr")` `r Citep(bib[["knitr"]])` and `r BiocStyle::CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r BiocStyle::CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`.

```{r vignetteBiblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
PrintBibliography(bib, .opts = list(hyperlink = "to.doc", style = "html"))
```
