---
title: "Computing Polygenic Score using GXwasR"
author: 
  - name: "Banabithi Bose"
    affiliation:
    - University of Colorado Anschutz Medical; 
    - Northwestern University
    email: banabithi.bose@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: false
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('GXwasR')`"
vignette: >
  %\VignetteIndexEntry{GXwasR_PGS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    fig.path = "./figures/",
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```


```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Bib setup
library("RefManageR")

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    GXwasR = citation("GXwasR")[1]
)
```

**This document provides a tutorial for computing state-of-the-art polygenic scores and sex-aware polygenic scores utilizing the GXwasR R package.**

```{r}
## Call some libraries (Users can ignore this.)
library(printr)
library(rmarkdown)
```

## Load the GXwasR and Example Datasets:

```{r}
## Load GXwasR
library(GXwasR)

## Load Example Data
data(Summary_Stat_Ex1)
data(Example_phenofile)
data(Example_covarfile)
data(Example_pthresoldfile)
data(highLD_hg19)
```

# Compute Standard PGSs using ComputePGS():

## About ComputePGS()

```{r}
help(ComputePGS, package = GXwasR)
```

## Running ComputePGS()

```{r, compute_pgs, fig.width=10,fig.height=6}
DataDir <- GXwasR:::GXwasR_data()
ResultDir <- tempdir()
finput <- "GXwasR_example"
summarystat <- Summary_Stat_Ex1[, c(2, 4, 7, 1, 3, 12)]
phenofile <- Example_phenofile # Cannot be NULL, the interested phenotype column should be labeled as "Pheno1".
covarfile <- Example_covarfile
clump_p1 <- 0.0001
clump_p2 <- 0.0001
clump_kb <- 500
clump_r2 <- 0.5
byCHR <- TRUE
pthreshold <- Example_pthresoldfile$Threshold
ld_prunning <- TRUE
highLD_regions <- highLD_hg19
window_size <- 50
step_size <- 5
r2_threshold <- 0.02
nPC <- 6 # We can incorporate PCs into our PGS analysis to account for population stratification.
pheno_type <- "binary"
ldclump <- FALSE
pheno_type <- "binary"
PGSresult <- ComputePGS(DataDir, ResultDir, finput, summarystat, phenofile, covarfile, effectsize = "BETA", LDreference = "GXwasR_example", ldclump = FALSE, clump_p1, clump_p2, clump_r2, clump_kb, byCHR = TRUE, pthreshold = pthreshold, highLD_regions = highLD_regions, ld_prunning = TRUE, window_size = 50, step_size = 5, r2_threshold = 0.02, nPC = 6, pheno_type = "binary")
PGSresult$PGS_plot
```

```{r}
## Table 1: A table showing 10 samples with phenotype, covariates, and sex-combined PGS scores.
PGS <- PGSresult$PGS
PGS[1:10, ]
## The best threshold
BestPvalue <- PGSresult$BestP$Threshold
BestPvalue
```

# Computing Sex-Aware PGS

## Datasets for computing PGS

**Discovery Data** i.e., GWAS summary statistics with mandatory columns such as "SNP"(SNP names), "A1"(effect allele), "BETA"(effect size in beta value) and "P"(p-value).

```{r}
# Example discovery data is included in GXwasR
data(Summary_Stat_Ex1)
summarystat <- Summary_Stat_Ex1[, c(2, 4, 7, 1, 3, 12)]
```

**Target Data** i.e., genotype dataset in plink .bed, .bim and .fam format.

```{r}
# Example target data is included in GXwasR
DataDir <- GXwasR:::GXwasR_data()
finput <- "GXwasR_example"
```

## Quality control of the datasets before computing sex-aware PGS.

Users must first ensure that both datasets  are mapped to the **same genome build**. The quality control procedure is:

**(A) Discovery Data: Remove multi-allelic, indels and ambiguous (A/T or C/G) SNPs. Then remove SNPs with minor allele frequency (MAF) < 0.05 and quality info score < 0.1.  For this filtering, R packages like data.table, dplyr, tidyverse can be used.**

**(B) Target Data: Remove multi-allelic, indels and ambiguous (A/T or C/G) SNPs. Then remove SNPs with MAF < 0.05 and Hardy Weinberg Equilibrium (hwe) > e-10. For this filtering steps, users should use `FilterAllele()` and `QCsnp()` functions in R.**

Let's see how to use these functions. For the details of these utility functions, please follow [https://boseb.github.io/GXwasR/articles/GXwasR_overview.html](https://boseb.github.io/GXwasR/articles/GXwasR_overview.html).

We will use `FilterAllele()` on target data to filter out the multi-allelic SNPs. Also, note that GXwasR does not work with multi-allelic variants. So, this filtering step is critical.

```{r}
# Target data
DataDir <- GXwasR:::GXwasR_data()
ResultDir <- tempdir()
finput <- "GXwasR_example"
foutput <- "filtered_multiallelic"
x <- FilterAllele(DataDir, ResultDir, finput, foutput)
```

We will next use `QCsnp()` to remove ambiguous (A/T or C/G) SNPs and SNPs with MAF < 0.05 and Hardy Weinberg Equilibrium (hwe) > e-10.

```{r}
## Since there was no multiallelic SNPs, we will continue with original input data.
finput <- "GXwasR_example"
foutput <- "filtered_step1"
geno <- NULL
maf <- 0.05
casecontrol <- FALSE
caldiffmiss <- FALSE
diffmissFilter <- FALSE
dmissX <- FALSE
dmissAutoY <- FALSE
monomorphicSNPs <- TRUE
ld_prunning <- FALSE
casecontrol <- FALSE ## Since the filtering doesn't require us to run on cases and controls separately, we will make this parameter FALSE.
hweCase <- NULL
hweControl <- NULL
hwe <- 1e-10
monomorphicSNPs <- FALSE
ld_prunning <- FALSE

x <- QCsnp(DataDir = DataDir, ResultDir = ResultDir, finput = finput, foutput = foutput, geno = geno, maf = maf, hweCase = hweCase, hweControl = hweControl, hwe = hwe, ld_prunning = ld_prunning, casecontrol = casecontrol, monomorphicSNPs = monomorphicSNPs, caldiffmiss = caldiffmiss, dmissX = dmissX, dmissAutoY = dmissAutoY, diffmissFilter = diffmissFilter)
```

**(C) Check whether SNPs present in the GWAS discovery dataset are in target dataset.**

  Gather SNPs common to discovery and target datasets.

```{r}
SNP1 <- unique(summarystat$SNP)
targetbim <- read.table(paste0(ResultDir, "/filtered_step1.bim"))
SNP2 <- unique(targetbim$V2)
commonSNP <- intersect(SNP1, SNP2) ## 991 SNPs are common between our discovery data and target data
```

## Filtering discovery data and target data to retain SNPs common to both datasets

```{r}
commonSNP <- data.table::as.data.table(commonSNP)
colnames(commonSNP) <- "SNP"
NewDiscoveryData <- merge(commonSNP, summarystat, by = "SNP")
## For making target data with common SNPs, we will use FilterSNP().
SNPvec <- commonSNP
# Need to copy filtered_step1 file from ResultDir to DataDir
ftemp <- list.files(paste0(ResultDir, "/"), pattern = "filtered_step1")
invisible(file.copy(paste0(ResultDir, "/", ftemp), DataDir))
finput <- "filtered_step1"
foutput <- "NewtargetData"
FilterSNP(DataDir, ResultDir, finput, foutput, SNPvec, extract = TRUE)
```

Now, discovery and target datasets are ready for computing sex-combined and sex-stratified PGS.

# Sex-combined PGS computation
**(A) Perform LD-clumping of the discovery data using the target data as the reference.**

**(B) Compute PGS at a variety of p-value thresholds.**

**(C) Select best threshold based on R-square (for quantitative trait) or MacFadden R-square (for binary trait) using generalized linear model with covariates, such as genetic PCs.**

Use `ComputePGS()` to perform the steps (A), (B) and (C).

```{r, compute_pgs_2, fig.width=10,fig.height=6}
# Running
# Filtered target data needs to be copied from ResultDir to DataDir.
ftemp <- list.files(paste0(ResultDir, "/"), pattern = "NewtargetData")
invisible(file.copy(paste0(ResultDir, "/", ftemp), DataDir))
finput <- "NewtargetData"

# Filtered discovery data.
# Need to maintain the first three coulmn of this dataset as SNP ID, Effect Allele and Effect Size
summarystat <- NewDiscoveryData
phenofile <- Example_phenofile # Cannot be NULL, the interested phenotype column should be labeled as "Pheno1".
## Added "AGE" and "testcovar" as covariates.
covarfile <- Example_covarfile
clump_p1 <- 0.0001
clump_p2 <- 0.0001
clump_kb <- 500
clump_r2 <- 0.5
byCHR <- TRUE
pthreshold <- Example_pthresoldfile$Threshold
ld_prunning <- TRUE
highLD_regions <- highLD_hg19
window_size <- 50
step_size <- 5
r2_threshold <- 0.02
nPC <- 6 # We can incorporate PCs into our PGS analysis to account for population stratification.
pheno_type <- "binary"
effectsize <- "BETA"
ldclump <- FALSE
PGSresult <- ComputePGS(DataDir, ResultDir, finput, summarystat, phenofile, covarfile, effectsize = "BETA", LDreference = "GXwasR_example", ldclump = FALSE, clump_p1, clump_p2, clump_r2, clump_kb, byCHR = TRUE, pthreshold = pthreshold, highLD_regions = highLD_regions, ld_prunning = TRUE, window_size = 50, step_size = 5, r2_threshold = 0.02, nPC = 6, pheno_type = "binary")
PGSresult$PGS_plot
```
Figure 2: PGS analysis. (A) Sex-combined PGS analysis showing model fit across p-value thresholds,  PGS distributions by: (B) sex, (C) case-control status, and PGS stratified by: case-control status within: (D) males, and (E) females.

**Note:** In real population, plot B) should show similar distribution of PGS scores in males and females. Slight difference in male vs female cases maybe possible.
```{r}
## Table 2: A table showing 10 samples with phenotype, covariates, and sex-combined PGS scores.
PGS <- PGSresult$PGS
PGS[1:10, ]
## The best threshold
BestPvalue <- PGSresult$BestP$Threshold
BestPvalue
```

**(D) Check whether there is a different liability threshold between sexes.**

To evaluate whether there is a sex-specific liability threshold, users can test whether the genetic burden (measured by PGS) differs between male and female cases. Under the liability threshold model, a sex difference would suggest that one sex requires a higher genetic burden to manifest the trait (i.e., a higher threshold), while the other requires fewer or smaller genetic burden factors.
["Khramtsova EA, Wilson MA, Martin J, Winham SJ, He KY, Davis LK, Stranger BE. Quality control and analytic best practices for testing genetic models of sex differences in large populations. Cell. 2023 May 11;186(10):2044-2061. doi: 10.1016/j.cell.2023.04.014. PMID: 37172561; PMCID: PMC10266536."]

Users can perform regression analyses across different PGS thresholds within cases only to assess the consistency of the sex effect. In this model, sex is the main predictor, with genetic principal components (PCs) and other relevant covariates included for adjustment:

**PGS ~ Sex + PCs + Other Covariates.**


This test can be implemented using SexRegress(), which evaluates whether male and female cases differ significantly in their polygenic burden.

# About SexRegress()

```{r}
help(SexRegress, package = GXwasR)
```

# Run SexRegress in cases()

In this case, the argument “famfile” should be used for the case-only file from the target data.

```{r, fig.width=10,fig.height=6}
# Running
# First, we need to make fdata object.
library(GXwasR)
famfile <- read.table(paste0(ResultDir, "/NewtargetData.fam"))[, c(1, 2, 5)]
colnames(famfile) <- c("FID", "IID", "Sex")
prefdata <- merge(famfile, PGS, by = c("IID", "FID"))

fdata <- prefdata[, c(13, 3:12)]
fdata$Sex <- as.factor(as.character(fdata$Sex))
response_index <- 1
regressor_index <- 2

x <- SexRegress(fdata, regressor_index, response_index)
x
```

In this example, the association between PGS and sex is not significant. Users can test across different p-value thresholds of PGS as mentioned in step (E).

## Sex-stratified PGS computation

The steps are:

**(A) Generate separate male and female discovery data by stratified GWAS.**

We will use SumstatMale and SumstatFemale for these.

**(B) Prepare separate male and female target datasets.**

For this, we will utilize `GetMFPlink()`.

```{r}
library(GXwasR)
## We will use NewtargetData as input for this function.
finput <- "NewtargetData"
foutput <- "maletarget"
sex <- "males"
x <- GetMFPlink(DataDir = DataDir, ResultDir = ResultDir, finput = finput, foutput = foutput, sex = sex, xplink = FALSE, autoplink = FALSE)

foutput <- "femaletarget"
sex <- "females"
x <- GetMFPlink(DataDir = DataDir, ResultDir = ResultDir, finput = finput, foutput = foutput, sex = sex, xplink = FALSE, autoplink = FALSE)
```

**(C) Perform LD clumping separately in both discovery data.**

**(D) Compute PGS at a variety of p-value thresholds, using male-only discovery data to compute scores for female-only target data and vice versa.**

**(E) Select best threshold based on R-square (for quantitative trait) or MacFadden R-square (for binary trait) using generalized linear model with covariates, such as genetic PCs.**

We will utilize the `ComputePGS()` to perform the steps (C), (D) and (E).

```{r, fig.width=10,fig.height=6}
# Running
# Male and female target datasets need to be copied from ResultDir to DataDir.
ftemp <- list.files(paste0(ResultDir, "/"), pattern = "target")
invisible(file.copy(paste0(ResultDir, "/", ftemp), DataDir))
```

# Computing PGS in female-only target data using male-only discovery data

Evaluate the transferability of genetic risk prediction across sexes and to test whether associations identified in one sex generalize to the other.

```{r, compute_pgs_female, fig.width=10,fig.height=6}
# DataDir <- GXwasR:::GXwasR_data()
ResultDir <- tempdir()
finput <- "femaletarget"
# Filtered discovery data.
# Need to maintain the first three coulmn of this dataset as SNP ID, Effect Allele and Effect Size
summarystat <- Summary_Stat_Ex1[, c(2, 4, 7, 1, 3, 12)]
## Making phenofile only with females
famfile <- read.table(paste0(ResultDir, "/femaletarget.fam"))[, c(1, 2, 5)]
colnames(famfile) <- c("FID", "IID", "Sex")
famfileF <- famfile[famfile$Sex == 2, ]
phenofile <- merge(famfileF, Example_phenofile, by = c("FID", "IID")) # Cannot be NULL, the interested phenotype column should be labeled as "Pheno1".
## Added "AGE" and "testcovar" as covariates.
## Making cpvarfile with females
covarfile <- merge(famfileF, Example_covarfile, by = c("FID", "IID"))
clump_p1 <- 0.0001
clump_p2 <- 0.0001
clump_kb <- 500
clump_r2 <- 0.5
byCHR <- TRUE
pthreshold <- Example_pthresoldfile$Threshold
ld_prunning <- TRUE
highLD_regions <- highLD_hg19
window_size <- 50
step_size <- 5
r2_threshold <- 0.02
nPC <- 6 # We can incorporate PCs into our PGS analysis to account for population stratification.
pheno_type <- "binary"
ldclump <- FALSE
effectsize <- "BETA"
PGSresultFemale <- ComputePGS(DataDir, ResultDir, finput, summarystat, phenofile, covarfile, effectsize = "BETA", LDreference = "GXwasR_example", ldclump = FALSE, clump_p1, clump_p2, clump_r2, clump_kb, byCHR = TRUE, pthreshold = pthreshold, highLD_regions = highLD_regions, ld_prunning = TRUE, window_size = 50, step_size = 5, r2_threshold = 0.02, nPC = 6, pheno_type = "binary")
PGSresultFemale$PGS_plot
```

Figure 3: Female-specific PGS analysis showing: (A) model fit across p-value thresholds, PGS distributions in: (B) females, and (C) stratified by case-control status within females.

```{r}
## Table 3: A table showing eight male samples with phenotype, covariates, and with female-specific PGS scores.
PGSfemale <- PGSresultFemale$PGS
PGSfemale[1:10, ]
## The best threshold
BestPvalue <- PGSresultFemale$BestP$Threshold
BestPvalue
```

# Compute PGS for male-only target data using female-only discovery data

```{r, compute_pgs_male, fig.width=10,fig.height=6}
library(GXwasR)
data(Summary_Stat_Ex2)
# DataDir <- system.file("extdata", package = "GXwasR")
ResultDir <- tempdir()
finput <- "maletarget"
# Filtered discovery data.
# Need to maintain the first three coulmn of this dataset as SNP ID, Effect Allele and Effect Size
summarystat <- Summary_Stat_Ex2[, c(2, 4, 7, 1, 3, 12)]
## Making phenofile only with females
famfile <- read.table(paste0(ResultDir, "/maletarget.fam"))[, c(1, 2, 5)]
colnames(famfile) <- c("FID", "IID", "Sex")
famfileM <- famfile[famfile$Sex == 1, ]
phenofile <- merge(famfileM, Example_phenofile, by = c("FID", "IID")) # Cannot be NULL, the interested phenotype column should be labeled as "Pheno1".
## Added "AGE" and "testcovar" as covariates.
## Making cpvarfile with females
covarfile <- merge(famfileM, Example_covarfile, by = c("FID", "IID"))
clump_p1 <- 0.0001
clump_p2 <- 0.0001
clump_kb <- 500
clump_r2 <- 0.5
byCHR <- TRUE
pthreshold <- Example_pthresoldfile$Threshold
ld_prunning <- TRUE
highLD_regions <- highLD_hg19
window_size <- 50
step_size <- 5
r2_threshold <- 0.02
nPC <- 6 # We can incorporate PCs into our PGS analysis to account for population stratification.
pheno_type <- "binary"
ldclump <- FALSE
effectsize <- "BETA"
PGSresultmale <- ComputePGS(DataDir, ResultDir, finput, summarystat, phenofile, covarfile, effectsize = "BETA", LDreference = "GXwasR_example", ldclump = FALSE, clump_p1 = clump_p1, clump_p2 = clump_p2, clump_r2 = clump_r2, clump_kb = clump_kb, byCHR = TRUE, pthreshold = pthreshold, highLD_regions = highLD_regions, ld_prunning = TRUE, window_size = 50, step_size = 5, r2_threshold = 0.02, nPC = 6, pheno_type = "binary")
PGSresultmale$PGS_plot
```

Figure 4: Male-specific PGS analysis showing: (A) model fit across p-value thresholds, PGS distributions in: (B) males, and (C) stratified by case-control status within males.

```{r}
## Table 4: A table showing eight female samples with phenotype, covariates and with male-specific PGS scores.
PGSmale <- PGSresultmale$PGS
PGSmale[1:10, ]
## The best threshold
BestPvalue <- PGSresultmale$BestP$Threshold
BestPvalue
```

## Citing `GXwasR`

We hope that `r BiocStyle::Biocpkg("GXwasR")` will be useful for your research. Please use the following information to cite the package and the overall approach. Thank you!

```{r "citation"}
## Citation info
citation("GXwasR")
```

# Reproducibility

The `r BiocStyle::Biocpkg("GXwasR")` package `r Citep(bib[["GXwasR"]])` was made possible thanks to:

* R `r Citep(bib[["R"]])`
* `r BiocStyle::Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
* `r BiocStyle::CRANpkg("knitr")` `r Citep(bib[["knitr"]])`
* `r BiocStyle::CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`
* `r BiocStyle::CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])`
* `r BiocStyle::CRANpkg("sessioninfo")` `r Citep(bib[["sessioninfo"]])`
* `r BiocStyle::CRANpkg("testthat")` `r Citep(bib[["testthat"]])`

This package was developed using `r BiocStyle::Biocpkg("biocthis")`.

`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```

# Bibliography

This vignette was generated using `r BiocStyle::Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
with `r BiocStyle::CRANpkg("knitr")` `r Citep(bib[["knitr"]])` and `r BiocStyle::CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r BiocStyle::CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`.

```{r vignetteBiblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
PrintBibliography(bib, .opts = list(hyperlink = "to.doc", style = "html"))
```
