---
title: "Tutorial for heritability estimate and computing genetic correlation in GXwasR R package."
author: "By Banabithi Bose"
date: "11/13/2023"
output:      
    BiocStyle::html_document:
        toc: true
        toc_depth: 3  
        number_sections: true
        theme: united 
r_packages: rmarkdown
vignette: >
  %\VignetteIndexEntry{Tutorial for heritability estimate and computing genetic correlation in GXwasR R package.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}{inputenc}
---
```{r global_options, include = FALSE}
knitr::opts_chunk$set(message=FALSE,
tidy.opts=list(width.cutoff=60))
```
**This document provides a tutorial for heritability estimate and computing genetic correlation utilizing GXwasR R package.**


```{r}
## Call some libraries
library(GXwasR)
library(printr)
library(rmarkdown)
## Load the example .Rda files
data("GXwasRData")
source("/projects/b1137/BBose/GXwasR/R/GXwasR_main_functions.R")
source("/projects/b1137/BBose/GXwasR/R/GXwasR_helper_functions.R")
```
# <a id = "EstimateHerit():"> </a>Computing heritability estimate using EstimateHerit():
```{r}
help(EstimateHerit, package = GXwasR)
```
## <a id = "Running EstimateHerit() with model = 'GREML'"> </a> Running EstimateHerit() with model = "GREML"
```{r}
# Running
DataDir = system.file("extdata", package = "GXwasR")
#ResultDir = tempdir()
ResultDir <- "/projects/b1137/BBose/Herit"
#DataDir <- ResultDir
finput <- "GXwasR_example"
ncores = 3
model = "GREML"
byCHR = FALSE
r2_LD = 0
LDSC_blocks = 20
REMLalgo = 0
nitr = 100
cat_covarfile = NULL
quant_covarfile = NULL
prevalance = 0.01
partGRM = FALSE
autosome = TRUE
Xsome = TRUE
nGRM = 3
cripticut = 0.025
minMAF = NULL
maxMAF = NULL
hg = "hg19"
byCHR = TRUE
PlotIndepSNP = TRUE
IndepSNP_window_size = 50
IndepSNP_step_size = 5
IndepSNP_r2_threshold = 0.02

# For saving the plot in .jpeg file, make plotjpeg = TRUE.

H2 <- EstimateHerit(DataDir = DataDir, ResultDir = ResultDir, finput = finput, summarystat = NULL, ncores = 5, model = model, byCHR = TRUE, r2_LD = 0, LDSC_blocks = 20,REMLalgo = 0, nitr = nitr, cat_covarfile = NULL, quant_covarfile = NULL,prevalance = 0.01, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3, computeGRM = TRUE, grmfile_name = NULL, cripticut = 0.025, minMAF = NULL, maxMAF = NULL,hg = "hg19",PlotIndepSNP = TRUE, IndepSNP_window_size = 50,IndepSNP_step_size = 5,IndepSNP_r2_threshold = 0.02,highLD_regions = highLD_hg19, plotjpeg = TRUE, plotname = "Heritability_Plots")

#Heritability
H2
```

## Running the EstimateHerit() by providing chromosome-wise GRM files in ResultDir
Set:

computeGRM = FALSE

grmfile_name = "GXwasR"...... This is the prefix. These files are created in the previous run in ResultDir. Users are encouraged to check that folder.


```{r}

H2 <- EstimateHerit(DataDir = DataDir, ResultDir = ResultDir, finput = finput, summarystat = NULL, ncores = 5, model = model, byCHR = TRUE, r2_LD = 0, LDSC_blocks = 20,REMLalgo = 0, nitr = nitr, cat_covarfile = NULL, quant_covarfile = NULL,prevalance = 0.01, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3, computeGRM = FALSE, grmfile_name = "GXwasR", cripticut = 0.025, minMAF = NULL, maxMAF = NULL,hg = "hg19",PlotIndepSNP = TRUE, IndepSNP_window_size = 50,IndepSNP_step_size = 5,IndepSNP_r2_threshold = 0.02,highLD_regions = highLD_hg19, plotjpeg = FALSE, plotname = "Heritability_Plots")

#Heritability
H2
```

## Running the EstimateHerit() for genome-wide heritability estimate by creating GRM files
Set:

computeGRM = TRUE

byCHR = FALSE

grmfile_name = NULL.......if computeGRM = TRUE the grmfile_name is by default NULL.

In genome-wide case, there no plot will be generated.

```{r}

H2 <- EstimateHerit(DataDir = DataDir, ResultDir = ResultDir, finput = finput, summarystat = NULL, ncores = 5, model = model, byCHR = FALSE, r2_LD = 0, LDSC_blocks = 20,REMLalgo = 0, nitr = nitr, cat_covarfile = NULL, quant_covarfile = NULL,prevalance = 0.01, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3, computeGRM = TRUE, grmfile_name = NULL, cripticut = 0.025, minMAF = NULL, maxMAF = NULL,hg = "hg19",PlotIndepSNP = TRUE, IndepSNP_window_size = 50,IndepSNP_step_size = 5,IndepSNP_r2_threshold = 0.02,highLD_regions = highLD_hg19, plotjpeg = FALSE, plotname = "Heritability_Plots")

#Heritability
H2
```

## Running the EstimateHerit() for genome-wide heritability estimate by utilizing the pre-computed GRM files in ResultDir
Set:

computeGRM = FALSE

byCHR = FALSE

grmfile_name = "GXwasR"

In genome-wide case, there no plot will be generated.

```{r}

H2 <- EstimateHerit(DataDir = DataDir, ResultDir = ResultDir, finput = finput, summarystat = NULL, ncores = 5, model = model, byCHR = FALSE, r2_LD = 0, LDSC_blocks = 20,REMLalgo = 0, nitr = nitr, cat_covarfile = NULL, quant_covarfile = NULL,prevalance = 0.01, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3, computeGRM = FALSE, grmfile_name = "GXwasR", cripticut = 0.025, minMAF = NULL, maxMAF = NULL,hg = "hg19",PlotIndepSNP = TRUE, IndepSNP_window_size = 50,IndepSNP_step_size = 5,IndepSNP_r2_threshold = 0.02,highLD_regions = highLD_hg19, plotjpeg = FALSE, plotname = "Heritability_Plots")

#Heritability
H2
```

## <a id = "Running EstimateHerit() with model = 'LDSC'"> </a> Running EstimateHerit() genomewide with model = "LDSC" 
```{r}
model <- "LDSC"
data("GXwasRData")
DataDir = system.file("extdata", package = "GXwasR")
ResultDir = tempdir()
finput <- "GXwasR_example"
test.sumstats <- na.omit(Summary_Stat_Ex1[Summary_Stat_Ex1$TEST=="ADD",c(1:4,6:8)])
colnames(test.sumstats) <- c("chr","rsid","pos","a1","n_eff","beta","beta_se")
summarystat = test.sumstats
highLD_regions = highLD_hg19

H2 <- EstimateHerit(DataDir = DataDir, ResultDir = ResultDir, finput = "GXwasR_example", summarystat = summarystat, ncores, model = model, byCHR = FALSE, r2_LD = 0, LDSC_blocks = 20,REMLalgo = 0, nitr = 100, cat_covarfile = NULL, quant_covarfile = NULL,prevalance = 0.01, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3,cripticut = 0.025, minMAF = NULL, maxMAF = NULL,hg = "hg19",PlotIndepSNP = TRUE, IndepSNP_window_size = 50,IndepSNP_step_size = 5,IndepSNP_r2_threshold = 0.02,highLD_regions = highLD_hg19)
#Heritability
H2
```

## <a id = "Running EstimateHerit() with model = 'LDSC'"> </a> Running EstimateHerit() chromosome-wise with model = "LDSC"
```{r}
model <- "LDSC"
data("GXwasRData")
DataDir = system.file("extdata", package = "GXwasR")
ResultDir = tempdir()
finput <- "GXwasR_example"
test.sumstats <- na.omit(Summary_Stat_Ex1[Summary_Stat_Ex1$TEST=="ADD",c(1:4,6:8)])
colnames(test.sumstats) <- c("chr","rsid","pos","a1","n_eff","beta","beta_se")
summarystat = test.sumstats
highLD_regions = highLD_hg19

H2 <- EstimateHerit(DataDir = DataDir, ResultDir = ResultDir, finput = "GXwasR_example", summarystat = summarystat, ncores, model = model, byCHR = TRUE, r2_LD = 0, LDSC_blocks = 20,REMLalgo = 0, nitr = 100, cat_covarfile = NULL, quant_covarfile = NULL,prevalance = 0.01, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3,cripticut = 0.025, minMAF = NULL, maxMAF = NULL,hg = "hg19",PlotIndepSNP = TRUE, IndepSNP_window_size = 50,IndepSNP_step_size = 5,IndepSNP_r2_threshold = 0.02,highLD_regions = highLD_hg19)

# Chromosome-wise Heritability Estimate
H2
```

# <a id = "GeneticCorrBT()"> </a> Computing genetic correlation between traits using GeneticCorrBT()
```{r}
help(GeneticCorrBT, package = GXwasR)
```
## <a id = "Running GeneticCorrBT()"> </a> Running GeneticCorrBT() by computing GRM genome-wide
```{r}

# Running
DataDir <- system.file("extdata", package = "GXwasR")
ResultDir <- "/projects/b1137/BBose/Herit"
#ResultDir <- tempdir()
finput <- "GXwasR_example"
byCHR = FALSE
REMLalgo = 0
nitr <- 3
ncores <- 3
#phenofile <- "GCphenofile.phen"
#load("/projects/b1137/BBose/GXwasR/data/Example_phenofile.Rda")
data("GXwasRData")
phenofile <- Example_phenofile #Cannot be NULL, the interested phenotype column should be labeled as
cat_covarfile <- NULL
quant_covarfile <- NULL
partGRM <- FALSE #Partition the GRM into m parts (by row),
autosome = TRUE
Xsome <- TRUE
cripticut = 0.025
minMAF <- 0.01 # if MAF filter apply
maxMAF <- 0.04
excludeResidual = TRUE
GC <- GeneticCorrBT(DataDir = DataDir, ResultDir = ResultDir, finput = finput, byCHR = byCHR,
                    REMLalgo = 0, nitr = nitr, phenofile = phenofile, cat_covarfile = NULL, quant_covarfile = NULL, computeGRM = TRUE, grmfile_name = NULL, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3,cripticut = 0.025, minMAF = NULL, maxMAF = NULL,excludeResidual = TRUE, ncores = ncores)
##Genetic correlation from the last iteration. In this example there was no rg value since the program stopped after nitr-1 = 2 iteration due to convergence issue.
GC
```

## Running GeneticCorrBT() using precomputed genome-wide GRM 
```{r}

byCHR = FALSE
computeGRM = FALSE
grmfile_name = "GXwasR"

GC <- GeneticCorrBT(DataDir = DataDir, ResultDir = ResultDir, finput = finput, byCHR = byCHR,
                    REMLalgo = 0, nitr = nitr, phenofile = phenofile, cat_covarfile = NULL, quant_covarfile = NULL, computeGRM = FALSE, grmfile_name = grmfile_name, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3,cripticut = 0.025, minMAF = NULL, maxMAF = NULL,excludeResidual = TRUE, ncores = ncores)
##Genetic correlation from the last iteration. In this example there was no rg value since the program stopped after nitr-1 = 2 iteration due to convergence issue.
GC
```

## Running GeneticCorrBT() by computing GRM chromosome-wise
```{r}
byCHR = TRUE
computeGRM = TRUE
grmfile_name = NULL

GC <- GeneticCorrBT(DataDir = DataDir, ResultDir = ResultDir, finput = finput, byCHR = byCHR,
                    REMLalgo = 0, nitr = nitr, phenofile = phenofile, cat_covarfile = NULL, quant_covarfile = NULL, computeGRM = computeGRM, grmfile_name = grmfile_name, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3,cripticut = 0.025, minMAF = NULL, maxMAF = NULL,excludeResidual = TRUE, ncores = ncores)
GC
```

## Running GeneticCorrBT() using chromosome-wide pre-computed GRM
```{r}
byCHR = TRUE
computeGRM = FALSE
grmfile_name = "GXwasR"

GC <- GeneticCorrBT(DataDir = DataDir, ResultDir = ResultDir, finput = finput, byCHR = byCHR,
                    REMLalgo = 0, nitr = nitr, phenofile = phenofile, cat_covarfile = NULL, quant_covarfile = NULL, computeGRM = computeGRM, grmfile_name = grmfile_name, partGRM = FALSE, autosome = TRUE, Xsome = TRUE, nGRM = 3,cripticut = 0.025, minMAF = NULL, maxMAF = NULL,excludeResidual = TRUE, ncores = ncores)
GC
```
